Categorical and gradient aspects of wordlikeness\thanks{Thanks to Gene Buckley, Charles Yang, and audiences at NELS 43 for helpful comments.}

### 2012-11-05 3.tex
Writing on chapter 2 primarily, also added the two "MSes" and new graphs for chapter 2.

\chapter{Static and derived Turkish phonotactics}
\label{turkish}

@@

Two less familiar notations are used in this chapter. First, directional application conditions are assumed, so as to derive the left-to-right, iterative properties of the harmony rules. Since it was first proposed by \citet{Johnson1972}, considerable evidence for directional application has been adduced (e.g., \citealt{Anderson1974}: chap.~9, \citealt{Gorman2012e}, \citealt{Howard1972}:65f., \citealt{Kavitskaya2008}, \citealt{Kaye1982}, \citealt{Kenstowicz1977}:189f., \citeyear{Kenstowicz1979}:319f., \citealt{Piggott1975}, \citealt{Sohn1971}; see \citealt{McCarthy2003b} and \citealt{Wolf2011b} for recent reviews.) Further, \citet{Johnson1972} and \citet{Kaplan1994} prove directional application and simultaneous application are equivalent in terms of formal learnability.

@@

\citet[212]{Anderson1974} and \citet{Iverson1978} argue that suffix harmony in disharmonic roots found in (\ref{turkexcept}a) requires the rule governing suffix harmony alternations to be distinguished from a sequence structure constraint governing root harmony. Since the rule and sequence structure constraint are otherwise identical, this constitutes an undesirable ``duplication'' in the sense of \citealt{Kisseberth1970b}. However, the theory of exceptionality proposed in \emph{SPE} can account for these facts without duplication \citep[197f.]{Zonneveld1978}.\footnote{\citet[29f.]{Kiparsky1968} discusses a parallel case in Finnish, and proposes a similar separation between the exception-filled sequence structure constraint and an exceptionless suffix harmony process. \citet[171f.]{Howard1972} correctly observes that this duplication is unnecessary.}

@@

\citeauthor{Anderson1974} also notes that roots which fail to undergo suffix harmony, like (\ref{turkexcept}b), may themselves be harmonic. He takes to be evidence for the necessity of duplication:

\begin{quote}
\ldots{}there are words which are exceptions to harmony across boundaries (e.g., \emph{kabahat} `fault', \emph{kabahattı} `his fault') but which are perfectly regular internally. Since the morpheme structure condition and the phonological rule in this case have distinct classes of exceptions, it is clear that they cannot be identified. \citep[289]{Anderson1974}
\end{quote}

@@

One crucial detail is missing from this analysis: \textsc{Backness Harmony} needs to be prevented from overwriting the [$+$\textsc{Back}] specification of disharmonic roots, either by non-derived environment blocking \citep{Mascaro1976} or structure preservation condition \citep{Kiparsky1985}. However, any condition which prevents \textsc{Backness Harmony} from overwriting underlying backness specifications will reintroduce the duplication of sequence structure and phonological generalizations; under this analysis, disharmonic roots are no longer exceptional, despite considerable evidence (reviewed below) that they are marked in Turkish.\footnote{On the other hand, it is possible to interpret the presence of a single backness specification per root as a sort of default. A precedent for this is the surface-oriented interpretation of the tonal Obligatory Contour Principle proposed by \citet[134]{Goldsmith1976} and \citet{Odden1986}, under which adjacent identical tones are automatically attributed to a single underlying tone. However, this is merely a notational variant of the rule exceptionality account in which [$+$\textsc{Backness Harmony}] is the default.} The the underspecification analysis is rejected here on these grounds.

@@

%Generative phonological theory at the time of \emph{SPE} lacked the representational vocabulary to directly distinguish between word-initial /bn/, which is disallowed, and /bn/ as a licit syllable contact cluster (e.g., \emph{o}[b.n]\emph{oxious}), a point first made by \citep{Hooper1973}. It is not the introduction of prosodic primitives into generative phonology that explains the contrast between \emph{blick} and \emph{bnick}, though, but rather the principle that surface forms must be syllabified \citep[e.g.,][63f.]{Kiparsky1982b} which \emph{bnick} fails. If, as is standardly assumed, phonology repairs unsyllabifiable outputs \citep[e.g.][]{Ito1989a,Noske1992}, then \emph{bnick} is an impossible output, doomed to be modified (perhaps to \emph{nick}; \citealt[][19f.]{Wolf2009}), and /bnɪk/ is an impossible UR by Stampean occultation. 
%A further widespread assumption is that static co-occurrence restrictions validated statistically will be reflected by wordlikeness judgements. This assumption is so deeply entrenched that a number of recent studies of phonotactic learning forgo wordlikeness judgements altogether and instead model the lexical statistics themselves (e.g., \citealt{Coetzee2008a} and \citealt{Anttila2008} on Muna and Arabic, \citealt[][385]{Hayes2008a} on Shona and Wargamay, \citealt{Brown2010} on Gitksan). 
%Since the null hypothesis does not countenance statistical significance as evidence for the synchronic reality of a static lexical pattern, such patterns do not adjudicate between the null hypothesis and the alternative, and another source of evidence is needed. In the remainder of Part I, I turn to data from wordlikeness judgement tasks for these purposes. 
%In a wordlikeness task, a speaker is presented with nonce words and asked to report such intuitions about the ``possibility'' of a word. 
%The alternative hypothesis, then, is that a statistically reliable co-occurrence restriction will be reflected in native speakers' wordlikeness judgements. 
%The data below, drawn from Turkish, shows that 
%This chapter takes the argument one step farther by showing a contrast between static and derived constraints on URs in Turkish. Whereas the derived constraint has a robust effect on native speakers' wordlikeness judgements, the static constraint, which is equally statistically valid, has no effect on these judgements. Anticipating the conclusion, the simplest explanation for the non-effect of this static constraint on wordlikeness judgements is that speakers do not internalize static co-occurrence restrictions at all.
%As a quick validation of this claim, consider the results reported by \citet{Albright2007} and discussed in the previous chapter. English speakers asked to rate non-words on a scale from 1-7, where 7 indicates ``most like English'', favor non-words [blʌs, blɑd] (average scores 4.67 and 5.13) over [bnʌs, bnɑd] (average scores 2.06, 2.00).
%This hypothesis is not ahdered to by early research on sequence structure constraints \citep[e.g.,][]{SPR,Chomsky1965,SPE,Stanley1967}. Work in this vein sought to use sequence structure constraints to compress the lexicon as much as possible. This work has a second interest, namely in ``compressing'' the lexicon as much as possible,
%motivations for the development of a theory of static co-occurrence restrictions as developed by \citet{SPR}, \citet{Stanley1967} and others is to account for speakers' abilities to distinguish between accidental and structural gaps. \citet{Chomsky1965} propose that a theory of phonology must account for speakers' abilities to distinguish between accidental and structural gaps in their language, citing a contrast between \emph{blick}, which is judged to be a possible but unattested word of English, and \emph{bnick}, which is judged to be impossible. \emph{SPE} (p.~3810f.) places this contrast in a system of a system of sequence structure rules that also are used to eliminate lexical redundancy. In the phonological theory of the era, there is no process that could generate the \emph{blick} $\sim$ \emph{bnick} contrast or numerous other lexical regularities
%Generative phonological theory at the time of \emph{SPE} lacked the representational vocabulary to directly distinguish between word-initial /bn/, which is disallowed, and /bn/ as a licit syllable contact cluster (e.g., \emph{o}[b.n]\emph{oxious}), a point first made by \citep{Hooper1973}. It is not the introduction of prosodic primitives into generative phonology that explains the contrast between \emph{blick} and \emph{bnick}, though, but rather the principle that surface forms must be syllabified \citep[e.g.,][63f.]{Kiparsky1982b} which \emph{bnick} fails. If, as is standardly assumed, phonology repairs unsyllabifiable outputs \citep[e.g.][]{Ito1989a,Noske1992}, then \emph{bnick} is an impossible output, doomed to be modified (perhaps to \emph{nick}; \citealt[][19f.]{Wolf2009}), and /bnɪk/ is an impossible UR by Stampean occultation. 
%Further, \citet[][528f.]{Halle1975} rejects his own principle that lexical entries should be free of all redundancies (see \citealt[][201]{Reiss2003a} and \citealt{Vaux2003} for further discussion). This rejected principle of a redundancy-free lexicon is one of the two principles that motivates Morpheme Structure Constraints; the other is the subject of this and the following chapter.
%\citet{Berent2007a}
%\citet{Berent2007b}
%\citet{Berent2008a}
% 598: Highly marked inputs are repaired in perception to abide by the grammatical restrictions of the language
% Exps 1-2: syllable count task (misperception due to linguistic experience; "disyllables were more likely to be perceived as monosyllables if their monosyllabic counterparts had a less marked cluster" in both Russian and English, not what OT predicts)
% Exps 3-4: AB task
% Exps 5-6: priming effect (non-epenthetic and epenethetic things represented the same)
% Berent and Lennertz 2007: (640: If the high rate of monosyllabic errors to unmarked onsets is only due to phonetic failures to encode the input, then it is puzzling why the same onsets also yield the highest rate of accurate response.)
% Berent et al. 2008a:
%\citet{Peperkamp2007}
% Peperkamp 2007: during one stage of processing phonology is undone with help from the lexicon...it only takes licit inputs and is irrelevant (p. 633)
%\citet{Dupoux1999}
% Dupoux et al. 1999: Japanese speakers still can't distinguish between ebzo and ebuzo in an ABX task using the SAME a or b.
%Once additional source of evidence on root (dis)harmony is inconclusive. There is a small class of bisyllabic words in which the second vowel, always [$+$\textsc{High}, $-$\textsc{Back}], alternates with zero. 

@@

%It is possible that \textsc{Backness Harmony} might produce a fluctuating \emph{ı} after root \emph{a}, but this does not obtain (\lastx b). However, this might simply indicate that the fluctuating vowel is epenthetic and that harmony applies before epenthesis (see \citealt{Clements1982} for both sides of this argument), making it less than a counterexample. 
%duplication problem: \citep{Kisseberth1970b} (see also \citealp[][401]{Stanley1967} and \emph{SPE}:382)
%(e.g., \emph{deve} `camel' vs. \emph{deva} `medicine', \emph{sene} `year' vs. \emph{sena} `praise').

@@

%  & {harf}         & {harfler}  & `(alphabetic) letter' \\ 
%  & {el}           & {eller}    & `hand'         \\
%  & {sap}          & {saplar}   & `stalk'        \\
%  & {son}          & {sonlar}   & `end'          \\
%b. & grip    & \alt{} & gırip    & `grippe'    \\ % unexpectedly back
%   & kredi   & \alt{} & kıredi   & `credit'    \\
%  & {el}           & {eli}          & {elin}         & `hand'    \\
%  & {yüz}          & {yüzü}         & {yüzün}        & `face'    \\
%  & {pul}          & {pulu}         & {pulun}        & `stamp'   \\

>>>>

\label{turkish}

@@

Two less familiar notations are used in this chapter. First, directional application conditions are assumed, so as to derive the left-to-right, iterative properties of the harmony rules. Since it was first proposed by \citet{Johnson1972}, considerable evidence for directional application has been adduced (e.g., \citealt{A74}: chap.~9, \citealt{Gorman2012e}, \citealt{Howard1972}:65f., \citealt{Kavitskaya2008}, \citealt{Kaye1982}, \citealt{KK77}:189f., \citeyear{KK79}:319f., \citealt{Piggott1975}, \citealt{Sohn1971}; see \citealt{McCarthy2003b} and \citealt{Wolf2011b} for recent reviews.) Further, \citet{Johnson1972} and \citet{Kaplan1994} prove directional application and simultaneous application are equivalent in terms of formal learnability.

@@

\citet[212]{A74} and \citet{Iverson1978} argue that suffix harmony in disharmonic roots found in (\ref{turkexcept}a) requires the rule governing suffix harmony alternations to be distinguished from a sequence structure constraint governing root harmony. Since the rule and sequence structure constraint are otherwise identical, this constitutes an undesirable ``duplication'' in the sense of \citealt{Kisseberth1970b}. However, the theory of exceptionality proposed in \emph{SPE} can account for these facts without duplication \citep[197f.]{Zonneveld1978}.\footnote{\citet[29f.]{Kiparsky1968} discusses a parallel case in Finnish, and proposes a similar separation between the exception-filled sequence structure constraint and an exceptionless suffix harmony process. \citet[171f.]{Howard1972} correctly observes that this duplication is unnecessary.}

@@

\citeauthor{A74} also notes that roots which fail to undergo suffix harmony, like (\ref{turkexcept}b), may themselves be harmonic. He takes to be evidence for the necessity of duplication:

\begin{quote}
\ldots{}there are words which are exceptions to harmony across boundaries (e.g., \emph{kabahat} `fault', \emph{kabahattı} `his fault') but which are perfectly regular internally. Since the morpheme structure condition and the phonological rule in this case have distinct classes of exceptions, it is clear that they cannot be identified. \citep[289]{A74}
\end{quote}

@@

One crucial detail is missing from this analysis: \textsc{Backness Harmony} needs to be prevented from overwriting the [$+$\textsc{Back}] specification of disharmonic roots, perhaps by structure preservation condition \citep{Kiparsky1985}. However, any condition which prevents \textsc{Backness Harmony} from overwriting underlying backness specifications will reintroduce the duplication of sequence structure and phonological generalizations; under this analysis, disharmonic roots are no longer exceptional, despite considerable evidence (reviewed below) that they are marked in Turkish.\footnote{On the other hand, it is possible to interpret the presence of a single backness specification per root as a sort of default. A precedent for this is the surface-oriented interpretation of the tonal Obligatory Contour Principle proposed by \citet[134]{Goldsmith1976} and \citet{Odden1986}, under which adjacent identical tones are automatically attributed to a single underlying tone. However, this is merely a notational variant of the rule exceptionality account in which [$+$\textsc{Backness Harmony}] is the default.} The the underspecification analysis is rejected here on these grounds.

@@

@@

@@

% 
%  & {harf}         & {harfler}  & `(alphabetic) letter' \\ 
%  & {el}           & {eller}    & `hand'         \\
%  & {sap}          & {saplar}   & `stalk'        \\
%  & {son}          & {sonlar}   & `end'          \\
%b. & grip    & \alt{} & gırip    & `grippe'    \\ % unexpectedly back
%   & kredi   & \alt{} & kıredi   & `credit'    \\
%  & {el}           & {eli}          & {elin}         & `hand'    \\
%  & {yüz}          & {yüzü}         & {yüzün}        & `face'    \\
%  & {pul}          & {pulu}         & {pulun}        & `stamp'   \\
